Finding and controling biases in dark matter halo concentration estimates

### 2016-07-08 paper/ms.tex
y

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title{Finding and controling biases in dark matter halo concentrations estimates}
\author{
  C. N. Poveda-Ruiz \thanks{cn.poveda542@uniandes.edu.co}$^{1}$
  J. E. Forero-Romero \thanks{je.forero@uniandes.edu.co}$^{1}$
  J. C. Mu\~noz-Cuartas \thanks{juan.munozc@udea.edu.co}$^{2}$
}

@@

\begin{abstract}
We use bootstrapping to estimate the bias of concentration
estimates on N-body dark matter halos as a function of particle
number. 
We find that algorithms based on the maximum radial velocity
and radial particle binning have increaing biases with decreasing 
particle numbers.
For instance, these methods overestimate concentrations by 
$15\%-20\%$ for halos sampled with $200$ particles.  
To overcome this bias we propose a new algorithm based on the
integrated mass profile. 
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.
This method reduces the bias to $2\%-3\%$ for halos sampled with
$200$ particles.  
To keep the biases of the velocity and density methods to the same
low levels as in the new method, dark matter halos should be sampled
with at least $\sim 4000$ particles. 
We also show that the mass-concentration relationship could be
shallower than expected from numerical experiments once the biases 
of the different concentration measurements are taken into account.
From these results we believe that bootstrapping and the
concentration estimation based on the integrated mass are
valuable tools to probe the internal structure of dark matter halos
in numerical simulations.
\end{abstract}

@@

In this paper we present a third alternative based on fitting the
integrated mass profile.  
This approach has two advantages with respect to the above mentioned methods.  
It does not involve any data binning and does not throw away data points.  
This translates into a robust estimate even at low resolution/particle numbers.  
Furthermore, since the method does not require any binning, there is no need to
tune numerical parameters.   
This method provides a new independent method to estimate the
concentration parameter.

In order to asses the quality of the three different methods we use
bootstraping on halos that are sampled with $10^{5}$ particles.
This allows us to estimate the bias and standard deviation on the
concentration estimates as a function of particle number.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{avg_foff_bolshoi.pdf}
  \includegraphics[width=0.48\textwidth]{sigma_foff_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the fractional offset ($f_{\rm
    off}$Eq. \ref{eq:f_off}) between the concentration in a halo at
  high resolution versus its downsampled versions.  Each point is the
  result of 100 downsampling iterations.  The velocity method
  noticeably overestimates the concentration up to a factor of $0.25$,
  while the new method only underestimates the concentrations by up to
  a factor of $0.05$.
Uncertainties on different concentration estimates.  The
  large symbols show the standard deviation of the fractional offset
  ($f_{\rm off}$ as defined in Eq. \ref{eq:f_off}) from the downsampling
  experiments.  The small dots correspond to the $1\sigma$
  uncertainty on the concentration for each halo as estimated by the
  MCMC algorithm used for the new method.
  The solid lines shows the $1/N$ tendency, which also corresponds
  to the particle number dependence of the $\langle|D|\rangle$ statistic
  (i.e. the accuracy to retrieve a known concentration in mock halos)
  of the new method.
  Uncertainties due to particle number fluctuations are always larger
  than the statistical uncertainties from the fitting process. 
    \label{fig:downsampling}}
\end{figure*}

@@

The new method uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).  We build it from N-body data following the
next steps.  First, we define the center of the halo to be at the
position of the particle with the lowest gravitational potential.
Then we rank the particles by their increasing radial distance from
the center.  From this ranked list of $i=1,N$ particles, the total
mass at a radius $r_i$ is $M_i=i\times m_p$, where $r_i$ is the
position of the $i$-th particle and $m_p$ is the mass of a single
computational particle.  In this process we discard the particle at
the center.

@@

\begin{equation}
\chi^2(c)= \sum_{i=1}^{N}[\log m_i - \log m(< x_i;c)]^2,
\end{equation}
%
where $m(<x_i;c)$ corresponds to the values in Eq.(\ref{eq:profile})
at $x=x_i$ for a given value of the concentration parameter $c$ and
the $i$ index sums over all the particles in the numerical profile.
From the $\chi^2$ distribution we find the optimal value of the
concentration and its associated uncertainty.

@@

We use data from the Bolshoi cosmological simulation that follows the
non-linear evolution of a dark matter density field sampled with
$2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.  
The cosmological parameters use a Hubble parameter $h=0.73$, a matter density
$\Omega_m=0.3071$ and a normalization of the power spectrum
$\sigma_8=0.82$. 
The
data is publicly available at \url{http://www.cosmosim.org/}.  More
details about the structure of the database and the simulation can be
found in \citep{2013AN....334..691R}.

@@

We use a set of two very different simulations to demonstrate that the 
results of our bootstraping tests are independent of the kind of
parent simulation.

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data.  All halos are
  used regardless of their relaxation state.  The lines correspond to
  the median concentration values in each mass bin.  The shaded region
  presents 10 and 90 per cent spread.  For clarity we only show the
  spread for the concentrations estimated using the new method. The
  other two methods have a similar spread. The dotted line corresponds
  to fits reported by \citep{Prada2012}.
  \label{fig:concentration}. The left panel shows the raw results
  coming from each algorithm. The right panel introduces a
  correction on the velocity and integrated mass results. 
  The corrections depends on the number of particles following the
  results of the downsampling experiments presented in Figure
  \ref{fig:downsampling}.} 
\end{figure*}

@@

We use bootstraping of individual high resolution halos to quantify
biases for different methods used to esteimate concentrations.

In this experiment we take halos sampled with at least $10^{5}$
particles and subsample each halo by factors of $2$ up to $10^{3}$ to
measure the concentration every time the halo is resampled.

The average concentration value for the largest number of particles,
$c_{N_{max}}$, provides with a baseline to compare all the other
results.  
We use the following statistic
\begin{equation}
f_{\rm off} = c_{N} / c_{N_{max}} - 1,
\label{eq:f_off}
\end{equation}
to account for the offset between the concentration at a given
downsampled particle number $c_{N}$ and the baseline $c_{N_{max}}$.

We select $8$ halos from the simulation that have at
least $3\times 10^5$ particles.  
We then randomly downsample the total number of particles in each halo
by factors of $10, 20, 45, 100, 215, 460, 1000$ to measure its
concentration with both the integrated mass and the velocity
method. For each downsampling factor we repeat the operation $100$
times.

Figure \ref{fig:downsampling} summarizes our results.  The plot shows
the average value of $f_{\rm off}$ as a function of the particle
number.  As expected for large enough sample particle numbers,
$N_{s}>5\times 10^3$, the results of the two algorithms coincide and
$\langle f_{\rm off}\rangle\sim 0$.  For lower number of particles,
the results of the two algorithms deviate from the expected
concentration value.  The velocity method overestimates by a factor
$\langle f_{\rm off}\rangle\sim 0.25$ the concentration values for the
lowest sample particle numbers, $N_{s}\sim 100$.  Around the same
sampling scale, the new algorithm shows a more stable behaviour with
an underestimate by a factor of $\langle f_{\rm off}\rangle \sim 0.02$.
This shows that in the more complex setup of a cosmological 
  simulation the new algorithm continues to be stable below the $5\%$
level for low particle numbers.

@@

Figure \ref{fig:downsampling_err} summarizes different uncertainty
results.  
Large symbols show the standard deviation on $f_{\rm off}$ from the
downsampling experiment.  
Small dots show the uncertainty on each individual concentration value
as estimated by the MCMC algorithm. 
The line shows the $\propto N^{-1}$ locus.  
This figure demonstrates that the statistical uncertainties are at
least one order of magnitude smaller than the uncertainty associated
to downsampling experiment.

>>>>

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title{Finding and controling biases in dark matter halo concentration estimates}
\author{
  C. N. Poveda-Ruiz \thanks{cn.poveda542@uniandes.edu.co}$^{1}$
  J. E. Forero-Romero \thanks{je.forero@uniandes.edu.co}$^{1}$
  J. C. Mu\~noz-Cuartas \thanks{juan.munozc@udea.edu.co}$^{2}$
}

@@

\begin{abstract}
We use bootstrapping to estimate the bias of concentration
estimates on N-body dark matter halos as a function of particle
number. 
We find that algorithms based on the maximum radial velocity
and radial particle binning have increaing biases with decreasing 
particle numbers.
These methods overestimate the concentration by 
$15\%-20\%$ for halos sampled with $200$ particles, and $7\%$-$8\%$
for halos sampled with $500$ particles. 
To control this bias level we propose a new algorithm based on the
integrated mass profile. 
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.
This method reduces the bias to $2\%-3\%$ for halos sampled with
$200$-$500$ particles.
To keep the biases of the velocity and density methods to the same
low levels as in the new method, only halos sampled
with at least $\sim 4000$ particles should be used.
We also show that the mass-concentration relationship could be
shallower than expected from numerical experiments once the biases 
of the different concentration measurements are taken into account.
From these results we believe that bootstrapping and the
concentration estimation based on the integrated mass are
valuable tools to probe the internal structure of dark matter halos
in numerical simulations.
\end{abstract}

@@

In this paper we use bootstraping to estimate the bias and standard
deviation on the concentration estimates as a function of particle
number. 
We show that the two standard methods to estimate concentrations have
increasing biases for decreasing particle numbers.

This motivates us to present a third alternative based on fitting the
integrated mass profile.  
This approach has two advantages with respect to the above mentioned methods.  
It does not involve any data binning and does not throw away data points.  
This translates into a robust estimate even at low resolution/particle numbers.  
Furthermore, since the method does not require any binning, there is no need to
tune numerical parameters.   
This method provides a new independent method to estimate the
concentration parameter.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{avg_foff_bolshoi.pdf}
  \includegraphics[width=0.48\textwidth]{sigma_foff_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{\emph{Left panel}. Average value of the fractional offset ($f_{\rm
    off}$ in Eq. \ref{eq:f_off}) between the concentration in a halo at
  high resolution versus its downsampled versions. 
  Thick lines
  correspond to massive halos in the Bolshoi simulation, thin lines to
  the Via Lactea simulation.
  For each case close to $10000$ bootstrapping samples are used.
  The velocity method noticeably overestimates the concentration up to
  a factor of $0.20$, 
  while the new method only underestimates the concentrations by up to
  a factor of $0.04$.
  \emph{Right panel.}
  1$\sigma$ uncertainties on the offset and the concentration
  estimates from the integrated mass method.  
  Lines show the width between the 14 and 86 percentiles
  of the $f_{\rm off}$ distribution at fixed particle number. 
  The lines include the results for the three methods: maximum
  velocity, density and integrated mass.
  Grey symbols correspond to the $1\sigma$ uncertainty on each of
  the halos in the Bolshoi Simulation as a result of the MCMC
  sampling. 
  To allow a fair comparison with $f_{\rm off}$, this uncertainty has
  been normalized by the preferred concentration value for each halo. 
  In both cases (bootstrapping for individul halos and MCMC fit of a
  halo population) the $1\sigma$ is of the same order with a
  decreasing behaviour as $1/\sqrt{N_s}$.
  \label{fig:downsampling}}
\end{figure*}

@@

Here we propose a new method to estimate the concentration. 
It uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).  We build it from N-body data following the
next steps.  First, we define the center of the halo to be at the
position of the particle with the lowest gravitational potential.
Then we rank the particles by their increasing radial distance from
the center.  From this ranked list of $i=1,N$ particles, the total
mass at a radius $r_i$ is $M_i=i\times m_p$, where $r_i$ is the
position of the $i$-th particle and $m_p$ is the mass of a single
computational particle.

@@

\begin{equation}
\chi^2(c)= \frac{1}{N}\sum_{i=1}^{N}[\log m_i - \log m(< x_i;c)]^2,
\end{equation}
%
where $m(<x_i;c)$ corresponds to the values in Eq.(\ref{eq:profile})
at $x=x_i$ for a given value of the concentration parameter $c$ and
the $i$ index sums over all the particles in the numerical profile.
From the $\chi^2$ distribution we find the optimal value of the
concentration and its associated uncertainty.

@@

We use two very different simulations to be sure that the results (bias,
uncertainty on the bias) measured using bootstraping tests are
independent of the kind of parent simulation.

We use data from the Bolshoi cosmological simulation that follows the
non-linear evolution of a dark matter density field sampled with
$2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.  
The cosmological parameters use a Hubble parameter $h=0.73$, a matter density
$\Omega_m=0.3071$ and a normalization of the power spectrum
$\sigma_8=0.82$. 
The data is publicly available at \url{http://www.cosmosim.org/}.  
Details about the structure of the database and the simulation can be
found in \citep{2013AN....334..691R}.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.49\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data.  
  We only used relaxed halos.
  All halos are
  The lines correspond to the median concentration values in each mass
  bin.  
  The shaded region presents 10 and 90 per cent spread. 
  For clarity we only show the spread for the concentrations estimated
  using the new method.  
  The other two methods have a similar spread. 
  The dotted line corresponds to fits reported by \citep{Prada2012}.
  \label{fig:concentration}. 
  The left panel shows the raw results coming from each algorithm. 
  The right panel introduces a correction on the velocity and
  integrated mass results following the results of the downsampling
  experiments presented in Figure 
  \ref{fig:downsampling}.} 
\end{figure*}

@@

In our bootstrapping experiments we take halos sampled with at least
$10^{5}$ particles and subsample each halo by factors of $2$ up to
$10^{3}$ to measure the concentration every time the halo is
resampled. 
Every time we keep fixed the virial radius and the center found for
the high resolution halo.
We extract in total 10000 bootstrapping samples for the Bolshoi
and the Via Lactea simulations.
In the case of the Bolshoi simulation we select 14 massive halos and
create 700 subsamples for each one. 
For the Via Lactea simulation only one halo is subsampled 10000
times.

The average concentration value for the largest number of particles,
$c_{N_{max}}$, provides with a baseline to compare all the other
results.  
We use the following statistic
\begin{equation}
f_{\rm off} = c_{N} / c_{N_{max}} - 1,
\label{eq:f_off}
\end{equation}
to account for the offset between the concentration at a given
downsampled particle number $c_{N}$ and the baseline $c_{N_{max}}$.
For each method we keep its corresponding concentration $c_{N_{max}}$
value at high resolution.

Figure \ref{fig:downsampling} summarizes our results.  
The plot on the left shows the average value of $f_{\rm off}$ as a
function of the particle number.  
As expected, for large enough sample particle numbers,
$N_{s}>4\times 10^3$, the results of the three algorithms coincide
below the $1\%$ level.
For a lower number of particles, the results deviate from the expected
concentration value.  
For $200$ particles the velocity method overestimates by a factor
$14\%$ while the density method overestimates it by $20\%$.
Around the same sampling scale, the new algorithm shows a more stable
behaviour underestimating the concentration only by a factor of
$1\%$-$2\%$

@@

The right panel in \ref{fig:downsampling} shows different uncertainty
results. 
The lines show the difference between the 14 and 86 percentiles in the
$f_{\rm off}$ distribution at fixed mass.
Each line corresponds to the three different methods to estimate the
concentration and the two simulations.
This shows that the bootstraping technique can help us to assign a
$1\sigma$ uncertainty to the concentration values at a fixed $N_s$. 
This uncertainty is shown as normalized by the preferred value of the
concentration, which in this case corresponds to $c_{N_{max}}$.  
The circles in the same figure show the $1\sigma$ uncertainty as
estimated from the MCMC chains using the integrated mass method on all
the relaxed halos in the Bolshoi simulation sample. 
To allow for a fair comparison with the bootstrapping results, this
uncertainty was normalized by the preferred concentration.
The MCMC uncertainties follow the same trend as the bootstraping
results. {\bf Hay algun significado profundo de esto o es lo que
  deberiamos esperar?}.

From these results we defined a universal $1\sigma$ fractional
uncertainty for the concentration as a function of the number of
particles $N_s$ as  
%
\begin{equation}
\sigma_{c} = \frac{0.40}{\sqrt{N_s/200}}.
\end{equation}
%
An interesting application of this result is that $\sigma_c=1$ for
$N_s=32$ particles, independent of the method used to estimate the
concentration.
